---
title: "Clustering"
author: "Vrinda Dhingra"
date: "2025-08-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading required libraries

```{r}

#Packages/Libraries to include

library(stats)
library(readxl)
library(nnls)
library(writexl)
library(CVXR)
library(gurobi) 
library(dplyr)
library(cluster)
library(dendextend)
library(urca) 

```


Loading the Price data for the index and its constituents. 
Calculating returns and defining the Asset Returns matrix and Index returns series.
Specifying the rolling window parameters. 
We choose one year in-sample and 3 months out-of-sample for our analysis. 


```{r}

#Data call

Price_data<-read_excel("......mention file path to load asset price data here....") 
Price_data<-as.matrix(Price_data)
Price_data<-Price_data[, -c(1)] # for removing the dates column 

price_row<-nrow(Price_data)-1
price_col<-ncol(Price_data)

num_price=as.numeric(Price_data)
Price.mat=matrix(c(num_price), nrow=price_row+1, ncol=price_col, byrow=F)

#Calculating Simple Returns

Ret_data<-matrix(nrow=price_row, ncol=price_col)
for(j in 1:price_col)
{
for(i in 1:price_row)
  {
    Ret_data[i,j]=(Price.mat[i+1,j]-Price.mat[i,j])/Price.mat[i,j]
  }
}

#Defining Index and Asset returns from Returns matrix

Ind_Ret<-Ret_data[, c(1)]
Ast_Ret<-Ret_data[, c(2:price_col)]

#Total number of assets
N=ncol(Ast_Ret)

#Total number of scenarios
T=nrow(Ast_Ret)

#cardinality and size bounds 
K=45

#Rolling_window strategy
in_sample=21*24
out_sample=21*3
roll=21*3
win_count=floor((T-in_sample-out_sample)/roll)+1

```


Declaring output matrices to store the portfolio weights from different models


```{r}

#OUTPUT MATRICES

W1=matrix(0, win_count, N) 
W2=matrix(0, win_count, N) 

```


```{r}

# MSE based tracking error minimization function to generate optimal portfolio after getting the selected assets from clustering 

mse_opt <- function(asset_ret, index_ret) {
  #' Minimize tracking error via quadratic programming.
  #'
  #' Solves:  min_w || index_ret – asset_ret %*% w ||^2
  #' Subject to:  sum(w) == 1,  w >= 0
  #'
  #' Parameters
  #' ----------
  #' asset_ret : matrix (T × N)
  #'   Matrix of asset returns (T time periods × N assets).
  #' index_ret : vector (T)
  #'   Vector of benchmark (index) returns.
  #'
  #' Returns
  #' -------
  #' numeric vector (N)
  #'   Optimal portfolio weights for N assets.
  #'
  #' Notes
  #' -----
  #' Uses CVXR to solve a convex quadratic program. If no solution is found,
  #' returns a zero-weight vector and issues a warning.
  # -----------------------------------------------------------------
  
  
  # Ensure input formats are numeric matrices/vectors
  asset_ret <- as.matrix(asset_ret)
  index_ret <- as.numeric(index_ret)
  
  # Number of assets
  N <- ncol(asset_ret)
  
  # Solve: min_w || index_ret - asset_ret %*% w ||^2
  # Subject to: sum(w) == 1, w >= 0
  
  library(CVXR)
  
  w <- Variable(N)
  objective <- Minimize(sum_squares(index_ret - asset_ret %*% w))
  constraints <- list(w >= 0, sum(w) == 1)
  
  problem <- Problem(objective, constraints)
  result <- solve(problem)
  
  if (result$status == "optimal") {
    return(as.numeric(result$getValue(w)))
  } else {
    warning(paste("Optimization failed with status:", result$status))
    return(rep(0, N))  # fallback: zero weights
  }
}


#Clustering:1 Cutting into K clusters and selecting one asset at random from each cluster


clust1<-function(AR, IR, L)
{
  #' Hierarchical clustering with equal representation.
  #'
  #' Forms L clusters of assets using correlation distance and complete linkage.
  #' From each cluster, one asset is randomly selected, ensuring equal representation across all clusters.
  #' Portfolio weights for the selected assets are then computed using mse_opt, and assigned back to the full set of assets.
  #' 
  #' Parameters
  #' ----------
  #' AR : matrix (T × N)
  #'   Asset returns.
  #' IR : vector (T)
  #'   Index returns.
  #' L : integer
  #'   Number of clusters to form.
  #'
  #' Returns
  #' -------
  #' numeric vector (N)
  #'   Portfolio weights across all assets, with non-zero weights only for
  #'   the assets selected from each cluster.
  # -----------------------------------------------------------------
  
returns_data=as.data.frame(AR)
returns_data<-returns_data + rnorm(nrow(returns_data), mean = 0, sd = 1e-6)

# Calculate correlation matrix
correlation_matrix<-cor(returns_data)

# Calculate distance matrix using correlation distance
distance_matrix<-as.dist(sqrt(2 * (1 - correlation_matrix)))

# Perform hierarchical clustering with complete linkage
clusters<-hclust(distance_matrix, method = "complete")

plot(clusters, hang=-1, cex=0.8, main="Hierarchical Clustering of SP500")

#Selecting assets from clusters

#Method 1: Random selection of one asset from each cluster 'K'

# Cut the dendrogram to get 250 clusters
cluster_cut1<-cutree(clusters, k=L)

# Output the clusters or use them for further analysis
table(cluster_cut1)

clust1_selection <- sapply(unique(cluster_cut1), function(cluster) {
  assets_in_cluster <- which(cluster_cut1 == cluster)
  sample(assets_in_cluster, 1)  # Randomly pick one asset from each cluster
})

index_returns <- as.matrix(IR)
asset_returns <- returns_data[, clust1_selection] 

weights<-mse_opt(asset_returns, index_returns)

total_assets <- ncol(returns_data)  # Total assets in the original dataset
final_portfolio_weights <- rep(0, total_assets)

# Assign the weights to the selected asset positions
final_portfolio_weights[clust1_selection] <- weights

# Display the final portfolio weights
return(final_portfolio_weights)
  
}


#Clustering method 2: Proportional selection of assets based on cluster size (size-weighted representation)


clust2<-function(AR, IR, K)
{
  #' Hierarchical clustering with proportional representation.
  #'
  #' Forms K clusters of assets using correlation distance and complete linkage.
  #' Instead of picking exactly one per cluster, this method selects a number of assets from each cluster in proportion to
  #' the cluster’s size relative to the total universe.
  #' Larger clusters contribute more representatives, smaller clusters fewer.
  #' Portfolio weights for the selected assets are then computed using mse_opt, and assigned back to the full set of assets.
  #' 
  #' Parameters
  #' ----------
  #' AR : matrix (T × N)
  #'   Asset returns.
  #' IR : vector (T)
  #'   Index returns.
  #' K : integer
  #'   Number of clusters to form.
  #'
  #' Returns
  #' -------
  #' numeric vector (N)
  #'   Portfolio weights across all assets, with non-zero weights only for
  #'   the proportionally selected assets.
  # -----------------------------------------------------------------

returns_data=as.data.frame(AR)
returns_data<-returns_data + rnorm(nrow(returns_data), mean = 0, sd = 1e-6)

# Calculate correlation matrix
correlation_matrix<-cor(returns_data)

# Calculate distance matrix using correlation distance
distance_matrix<-as.dist(sqrt(2 * (1 - correlation_matrix)))

# Perform hierarchical clustering with complete linkage
clusters<-hclust(distance_matrix, method = "complete")

plot(clusters, hang=-1, cex=0.8, main="Hierarchical Clustering of SP500")

#Selecting assets from clusters

# Cut the dendrogram to get 150 clusters
cluster_cut2<-cutree(clusters, k=K)

# Output the clusters or use them for further analysis
table(cluster_cut2)

total_assets <- length(cluster_cut2)

clust2_selection <- unlist(sapply(unique(cluster_cut2), function(cluster){
  assets_in_cluster <- which(cluster_cut2 == cluster)
  num_assets <- round((length(assets_in_cluster) / total_assets) * K)
  sample(assets_in_cluster, num_assets)
}))

index_returns <- as.matrix(IR)
asset_returns <- returns_data[, clust2_selection] 

weights<-mse_opt(asset_returns, index_returns)

total_assets <- ncol(returns_data)  # Total assets in the original dataset
final_portfolio_weights <- rep(0, total_assets)

# Assign the weights to the selected asset positions
final_portfolio_weights[clust2_selection] <- weights

# Display the final portfolio weights
return(final_portfolio_weights)
  
}

```


Main Script- Rolling window methodology to obtain optimal portfolios from clustering


```{r}

start=Sys.time()

for(i in 1:win_count)
{
  q=in_sample
  m1=1+roll*(i-1)
  m2=q+roll*(i-1)
  
  IR=Ind_Ret[c(m1:m2)]
  AR=Ast_Ret[c(m1:m2), c(1:N)]
  
  #Calculating portfolio weights
  
  W1[i,]=clust1(AR,IR,K)
  W2[i,]=clust2(AR,IR,K)
  
}

end=Sys.time()
total=end-start
total
  
```


Saving the final portfolio weights as Excel files


```{r}

df1=as.data.frame(W1)
df2=as.data.frame(W2)
  
write_xlsx(df1, ".....mention file path to store the weights file.../W_Clust1.xlsx")
write_xlsx(df2, ".....mention file path to store the weights file.../W_Clust2.xlsx")
  

```


