---
title: "Main File for Optimization Models"
author: "Vrinda Dhingra"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Calling the required libraries


```{r}
#Packages/Libraries to include

library(stats)
library(readxl)
library(writexl)
library(CVXR)
library(gurobi) 
library(nnls)
library(quantreg)
```

Next chunk:

1) Loading the Price data for the index and its constituents. Enter the path for the price data file in Price_data
2) Calculating returns (we calculate simple returns, log returns can also be used) 
3) Defining the Asset Returns matrix and Index returns series.
4) Specifying the rolling window parameters. 
   (We choose one year in-sample and 3 months out-of-sample for our analysis) 


```{r}

#Data call

Price_data<-read_excel("......enter your file path name here.......") 
Price_data<-as.matrix(Price_data)
Price_data<-Price_data[, -c(1)] # for removing the dates column 

price_row<-nrow(Price_data)-1
price_col<-ncol(Price_data)

num_price=as.numeric(Price_data)
Price.mat=matrix(c(num_price), nrow=price_row+1, ncol=price_col, byrow=F)

#Calculating Simple Returns

Ret_data<-matrix(nrow=price_row, ncol=price_col)
for(j in 1:price_col)
{
for(i in 1:price_row)
  {
    Ret_data[i,j]=(Price.mat[i+1,j]-Price.mat[i,j])/Price.mat[i,j]
  }
}

#Defining Index and Asset returns from Returns matrix

Ind_Ret<-Ret_data[, c(1)]
Ast_Ret<-Ret_data[, c(2:price_col)]

#Total number of assets
N=ncol(Ast_Ret)

#Total number of scenarios
T=nrow(Ast_Ret)

#cardinality specifications
K=45

#Model specific parameters
alpha=c(0.9, 0.75, 0.5, 0.1, 0.01)
delta=0.5
tau=0.5

#Rolling_window parameters
in_sample=21*24
out_sample=21*3
roll=21*3
win_count=floor((T-in_sample-out_sample)/roll)+1

#Portfolio weights output matrix
W=matrix(nrow=win_count, ncol=N)

```

Next chunk:

Main File: Rolling window framework
Here, call the source file (model/function) you want to work with. 

Example source file call:
For example, if you want to run the Index tracking optimization model with tracking error as MSE, then call 

source("MSE.R")



```{r}

source("MSE.R")

start_time=Sys.time()

for(i in 1:win_count)
{
  start=Sys.time()
  
  m1=1+roll*(i-1)
  m2=in_sample+roll*(i-1)
  
  IR=Ind_Ret[c(m1:m2)]
  AR=Ast_Ret[c(m1:m2), c(1:N)]
  
  #Calculating portfolio weights using the function call and passing required parameters

  W[i,]=mse(AR, IR, K)
  
  end=Sys.time()
  total=end-start
  total
  
  print(paste("Time taken for window", i, ":", total))
  
}

end_time=Sys.time()
total_time=end_time-start_time
total_time

print(paste("Time taken for the program:", total_time))

```


Last: Save the portfolio weights file as an excel sheet



```{r}

df_w=as.data.frame(W)
write_xlsx(df_w, ".....enter path where file is to be stored...")

```